Manuscript progress report

### 2023-12-04 Docs/ProgressReportAppl.Rmd
small sample data application

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
theme_set(theme_minimal())
```

@@

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))
```

@@

- 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction

@@

train_id <- sample(unique(df$id), size = N*0.8)
test_id <- setdiff(unique(df$id), train_id)

train_df <- df %>% filter(id %in% train_id)
test_df <- df %>% filter(id %in% test_id)
```

@@

mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=length(train_id), 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time
# dim(mat_est_unique)

@@

```{r}
plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
```

plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
```

```{r}
# plot correlation matrix
heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
```

@@

Becuase I would like to focus this report on subjects with failed Laplace approximation, I will not do these two things for now.

@@

follow up on the last report, I'll add some details here.

Let's set:

- $s$: bin index
- $t_{m_s}$: midpoint time of bin s
- $\mathscr{T}_s$: all observation points in bin s
- $Y_i^s$ all observations in bin s from subject i. $Y_i^S := \{Y_i(t_j), t_j \in \mathscr{T}_s\}$

Let's assume we have a new subject $u$ with maximum observation time $T_u$. Then the log-likelihood of this new subject would be:

@@

## One subject example

Now I am going to use one subject as an example to go over the approximation procedure:
```{r, class.source = "fold-show"}
## model
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

  # log-prior
  xi.prior <- dmvnorm(xi, mean = rep(0, Data$J), sigma=Data$tao, log = TRUE)

  # log-posterior likelihood
  eta <- Data$f0+Data$X %*% xi
  p <- exp(eta)/(1+exp(eta))
  LL <- sum(dbinom(x=Data$y, size = Data$n, prob=p, log = TRUE)) # log likelihood of Y|xi
  LP <- LL+sum(xi.prior) # unnormalized joint log likelihood of (Y, xi)

  # output
  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LP,
                   yhat=Data$y, parm=parm)
  return(Modelout)
}
```

@@

```{r, class.source = "fold-show"}
rand_id <- test_id[15]
df_new <- df %>% filter(id == rand_id & sind <= 540) # observation until 9am

# put data into correct format
ns <- as.vector(table(df_new$bin)) # number of observations
hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% select(Y) %>% unlist()# number of success
nf <- ns-hs # number of failure
max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)

# extract estimates from FPCA model to be used in prior distribution
tao <- diag(fpca_mod$evalues[1:K]) # variance of xi

# extract estimates from FPCA model to be used in posterior distribution
f0 <- fpca_mod$mu[1:max_bin]
N <- nrow(df_new2) # "sample size" which is in fact number of observed bins in this case for a new subject
n <- df_new2$ns # number of experiements at each bin
y <- df_new2$hs # outcome, number of 1
X <- fpca_mod$efunctions[1:max_bin, 1:K] # eigen/PC functions
J <- K # number of parameters/scores
mon.names <- "LP"

# parameter names
name_lst <- as.list(rep(0, J))
names(name_lst) <- paste("xi", 1:J, sep = "")
parm.names <- as.parm.names(name_lst) # names of parameters to estimate
pos.xi <- grep("xi", parm.names)

# data
MyData <- list(J=J, X=X, mon.names=mon.names,
                 parm.names=parm.names, pos.xi=pos.xi, y=y, n=n, tao=tao, f0=f0)
```

```{r, class.source = "fold-show"}
# fit laplace approximation
# initial values generated by GIV function in the same package
Fit <- LaplaceApproximation(Model, Data=MyData)
```

- Summarizing output

```{r, class.source = "fold-show", fig.height=8, fig.width=8}
# summarizing output
print(Fit)
plot(Fit, MyData)
```

```{r}
# prediction
eta_pred_out <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%Fit$Summary1[, "Mode"]

data.frame(bin = mid, eta_pred = eta_pred_out) %>% 
  right_join(df %>% filter(id==rand_id) %>%  select(Y, sind, bin), by = "bin") %>%
  ggplot()+
  geom_point(aes(x=sind, y = Y))+
  geom_line(aes(x=sind, y = exp(eta_pred)/(1+exp(eta_pred))))+
  labs(title = rand_id)

@@

## Full data application output

### fGFPCA

```{r}
# results generated from DataApplNHANES.R
load(here("Data/ApplOutput_fGFPCA.RData"))
```

```{r message=FALSE, warning=FALSE}
df_test_full <- df %>% 
  filter(id %in% test_id) %>%
  # filter(!id %in% skip_id) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))

rand_id2 <- sample(unique(df_test$id), 4)
# without interpolation, assume constant latent function value in each bin
df_test_full%>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540, col = "9am"))+
    geom_line(aes(x=bin, y = pred_t780, col = "1pm"))+
    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"))+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)",
         title = "Constant latent function within bin")
```

<!-- Calculate AUC -->

```{r}
# break by prediction window
lst_auc <- df_test_full %>% 
  mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
                      labels = c("0-9am", "9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
lst_auc <- split(lst_auc, f=lst_auc$window)
```

```{r}
# no interpolation
# up to 9am
auc_t540 <- lst_auc[2:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t540, x$Y), measure = "auc"))
auc_t540 <- lapply(auc_t540, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1pm
auc_t780 <- lst_auc[3:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t780, x$Y), measure = "auc"))
auc_t780 <- lapply(auc_t780, function(x){x@y.values[[1]]}) %>% unlist()

# up to 5pm
auc_t1020 <- lst_auc[4] %>% 
  lapply(function(x)performance(prediction(x$pred_t1020, x$Y), measure = "auc"))
auc_t1020 <- lapply(auc_t1020, function(x){x@y.values[[1]]}) %>% unlist()
```

## GLMMadaptvie

```{r}
load(here("Data/ApplOutput_GLMMadaptive.RData"))
```

- `r round(t_est_adglmm, 2)` minutes on model fitting and `r round(t_pred_adglmm, 2)` minutes on out-of-sample prediction.

- Below is the model fit on the training set

```{r}
summary(adglmm_mod)
```

```{r}
df_test_full2 <- df_test_full %>%
  filter(id %in% rand_id2)%>%
  select(id, sind, Y)

df_test_full2$pred_t540 <- df_test_full2$pred_t780 <- df_test_full2$pred_t1020 <- NA 
for(i in rand_id2){
  df_test_full2[df_test_full2$id==i & df_test_full2$sind>540, "pred_t540"]<- adglmm_pred_t540$newdata2 %>% filter(id ==i) %>% select(pred)

  df_test_full2[df_test_full2$id==i & df_test_full2$sind>780, "pred_t780"]<- adglmm_pred_t780$newdata2 %>% filter(id ==i) %>% select(pred)

  df_test_full2[df_test_full2$id==i & df_test_full2$sind>1020, "pred_t1020"]<- adglmm_pred_t1020$newdata2 %>% filter(id ==i) %>% select(pred)
}

# colSums(is.na(df_test_full2))
```

- Below is an example of predicted probablity on four random subjects

```{r, warning=FALSE}
df_test_full2 %>% 
  mutate_at(vars(pred_t540, pred_t780, pred_t1020), 
            function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
    geom_line(aes(x=sind, y = pred_t540, col = "9am"))+
    geom_line(aes(x=sind, y = pred_t780, col = "1pm"))+
    geom_line(aes(x=sind, y = pred_t1020, col = "5pm"))+
    geom_point(aes(x=sind, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

<!-- Calculate AUC  -->

```{r}
# up to 540
J <- 1440
auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
  # filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(540, 780, 1020, 1260, 1440), 
                      labels = c("9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t540_adglmm <- split(auc_t540_adglmm, f=auc_t540_adglmm$window)
auc_t540_adglmm <- auc_t540_adglmm[1:3] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t540_adglmm <- lapply(auc_t540_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 780
auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
  # filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(780, 1020, 1260, 1440), 
                      labels = c("1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t780_adglmm <- split(auc_t780_adglmm, f=auc_t780_adglmm$window)
auc_t780_adglmm <- auc_t780_adglmm[1:2] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t780_adglmm <- lapply(auc_t780_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1020
auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
  # filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(1020, 1260, 1440), 
                      labels = c("5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t1020_adglmm <- split(auc_t1020_adglmm, f=auc_t1020_adglmm$window)
auc_t1020_adglmm <- auc_t1020_adglmm[1] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t1020_adglmm <- lapply(auc_t1020_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

### AUC

- To use the same sample for comparison, I'd exclude the subjects with failed Laplace approximation from the test set for both methods

```{r}
options(knitr.kable.NA = '')
tb_auc <- data.frame(auc_t540, 
           c(NA, auc_t780), 
           c(NA, NA, auc_t1020),

           auc_t540_adglmm,
           c(NA, auc_t780_adglmm), 
           c(NA, NA, auc_t1020_adglmm))

colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)
```

```{r}
# time
t_fit <- c(t_local_glmm+t_fpca, t_est_adglmm)
units(t_fit) <- "mins"

t_test <- c(t_pred, t_pred_adglmm)
# units(t_pred)

t_total <- t_fit+t_test

idx1 <- idx2 <- idx3 <- c(1, 3, 3)
names(idx1) <- c("Total time", as.character(round(t_total, 2)))
names(idx2) <- c("Time on prediction", as.character(round(t_test, 2)))
names(idx3) <- c("Time on model fitting", as.character(round(t_fit, 2)))

t_fit <- as.character(round(t_fit, 2))
t_test <- as.character(round(t_test, 2))
t_total <- as.character(round(t_total, 2))

```{r}
tb_auc %>%
  kable(digit = 4, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(header=idx1) %>%
  add_header_above(header=idx2) %>%
  add_header_above(header=idx3) %>%
  add_header_above(c(" "=1, "fGFPCA"=3, "GLMMadaptive"=3))
```

- fGFPCA does not always have higher AUC, but definitely most of the time. 
- In fact there are two scenarios where GLMMadaptive outperforms fGFPCA slightly: 1) given 9am to predict 1-5pm; and 2) given 1pm and predict 5-9pm. Both of them have a 4-hour interval apart from the maximum observation time and and prediction window. Could it have anything to do the cyclic trend in the data? Is 4h the length of a cycle? I tried to look at it but really couldn't see anything...there are too many points lumped together!

```{r}
df %>% filter(id %in% rand_id2) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.2)+
  geom_vline(xintercept = c(540, 780, 1020, 1260))+
  scale_x_continuous(breaks = c(540, 780, 1020, 1260),
                     labels = c("9am", "1pm", "5pm", "9pm"))+
  facet_wrap(~id)
```

## Failed Laplace approximation

- This problem seems to be solved after changing the optimizer? (Method = "BFGS")
- The new optimizer also took less time
- The old one was SPG (Spectral projected gradient), which is a non-monotone line search. 
- The new one is BFGS (quasi-newton, the same in glmmTMB)

## Discussion

- What would be the reference method to compare our performance to? Original GLMMadaptvie perhaps is feasible on the entire dataset. But what should we use if we wanna do sub-sample comparison? Should we even do that for this data application (or simulation alone)? 
<!-- - There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects.  -->
<!-- - Is interpolation really necessary? It does not improve perfor- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...

>>>>

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
theme_set(theme_minimal())
```

@@

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutsampBayes.R"))
```

@@

- Original sample size took too much time in the debias step of fGFPCA method. 
- In fact, the debais step does not work very well in this case
- 500 subjects for training, 500 for out-of-sample prediction

@@

Ntr <- Nte <- 500

train_id <- unique(df$id)[1:500]
test_id <- unique(df$id)[501:1000]

train_df <- df %>% filter(id %in% train_id)
test_df <- df %>% filter(id %in% test_id)
# 1440*500
```

@@

mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=Ntr, 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time
# dim(mat_est_unique)

@@

```{r, eval=FALSE}
# plot correlation matrix
heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
```

@@

```{r projection}
# Re-evaluation
## grid extension
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
                   outer.ok = TRUE)$design  # evaluate B-splines on original grid

df_phi <- matrix(NA, J, K) 
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  df_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid
```

plot(mid, fpca_mod$efunctions[, 1], pch=16, xlab="bin", ylab="PC1", col = "blue",
     cex = 0.5)
lines(1:J, df_phi[, 1])

plot(mid, fpca_mod$efunctions[, 2], pch=16, xlab="bin", ylab="PC2", col = "blue",
     cex = 0.5)
lines(1:J, df_phi[, 2])

plot(mid, fpca_mod$efunctions[, 3], pch=16, xlab="bin", ylab="PC3", col = "blue",
     cex = 0.5)
lines(1:J, df_phi[, 3])

plot(mid, fpca_mod$efunctions[, 4], pch=16, xlab="bin", ylab="PC4", col = "blue",
     cex = 0.5)
lines(1:J, df_phi[, 4])
```

```{r revaluation}
df_phi <- data.frame(sind = 1:J, df_phi)
colnames(df_phi) <- c("sind", paste0("phi", 1:K))
train_df <- train_df %>% 
  left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)
# class(train_df$id)
# train_df$id2 <- factor(train_df$id, levels = train_id, labels = 1:length(train_id))
# train_df$id2 <- as.numeric(train_df$id2)
# range(unique(train_df$id2))

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t2-t1
```

```{r}
# mean
new_mu <- predict(debias_glmm, type = "terms")[1:J, 1] # extract re-evaluated mean

plot(mid, fpca_mod$mu, pch=16, xlab="bin", ylab="PC4", col = "blue",
     cex = 0.5)
lines(1:J, new_mu)

# eigenvalues
fpca_mod$evalues[1:K] # original
lambda_scaled <- 1/debias_glmm$sp[2:5] # extract re-evaluated lambda
fpca_mod$evalues[1:K]/lambda_scaled # ratio
```

```{r rescale}
## rescale
new_phi <- df_phi %>% select(starts_with("phi"))*sqrt(n_bin)
new_phi <- as.matrix(new_phi)

lambda_scaled <- lambda_scaled/n_bin
```

@@

@@

@@

```{r}
window <- seq(0, J, by = 8*60) # midnight,  8am, 4pm, midnight

test_df <- rename(test_df, t = sind)

converge_state_m <- matrix(NA, nrow = length(test_id), 2)
pred_list_m <- list()

t1 <- Sys.time()
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])
  # per max obs time
  for(tmax in window[2:3]){
    df_it <- df_i %>% filter(t <= tmax)
    max_rid <- nrow(df_it)

    # into a list
    MyData <- list(K=K, 
                   X=new_phi[1:max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(lambda_scaled), f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, 
                                parm = rep(0, K), 
                                Data=MyData, 
                                Method = "NM", 
                                Iterations = 1000,
                                CovEst = "Identity")
    converge_state_m[i, which(window[2:3]==tmax)] <- Fit$Converged
    score <- Fit$Summary1[, "Mode"]

    # prediction
    eta_pred_out <- new_mu+new_phi%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

  pred_list_m[[i]] <- df_i
}
t2 <- Sys.time()
t_pred <- t2-t1 # About 3.5 minutes

# mean(converge_state_m) # convergence rate 100%
```

```{r}
# check results
df_pred <- bind_rows(pred_list_m)
df_pred$pred480[df_pred$t<480] <- NA
df_pred$pred960[df_pred$t<960] <- NA
```

# GLMMadaptvie

```{r}
t1 <- Sys.time()
adglmm_mod <- mixed_model(Y ~ t, random = ~ t | id, 
                          data = train_df %>% mutate(t=sind/J),
                      family = binomial())
t2 <- Sys.time()
t_est_adglmm <- t2-t1 # model fitting took 20.82 mins
summary(adglmm_mod)
```

```{r}
test_df2 <- test_df %>% mutate(t=t/J)
pred_list_m2 <- list(
  predict(adglmm_mod, newdata = test_df2 %>% filter(t <= 480/J),
          newdata2 =  test_df2 %>% filter(t>480/J), 
          type = "subject_specific", type_pred = "link",
          se.fit = FALSE, return_newdata = TRUE)$newdata2,

  predict(adglmm_mod, newdata = test_df2 %>% filter(t <= 960/J),
          newdata2 =  test_df2 %>% filter(t>960/J), 
          type = "subject_specific", type_pred = "link",
          se.fit = FALSE, return_newdata = TRUE)$newdata2
)
```

```{r}
# check results
pred_list_m2 <- lapply(pred_list_m2, function(x){x %>% select(id, t, pred)})
df_pred2 <- test_df2 %>% 
  left_join(pred_list_m2[[1]], by = c("id", "t")) %>% 
  rename(pred480 = pred) %>% 
  left_join(pred_list_m2[[2]], by = c("id", "t")) %>% 
  rename(pred960 = pred) %>%
  mutate(t=t*J)
```

# Results

```{r}
rand_id_test <- sample(unique(test_id), 4)
ggarrange(
  df_pred %>% 
    filter(id %in% rand_id_test) %>%
    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
    ggplot()+
    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "fGFPCA"),

  df_pred2 %>% 
    filter(id %in% rand_id_test) %>%
    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
    ggplot()+
    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "GLMMadaptive"),
  nrow = 1, common.legend = T)

@@

```{r auc_func}
## a function to calculate AUC
get_auc <- function(y, pred){
  if(sum(is.na(y))>0 | sum(is.na(pred))>0){
    auc <- NA
  }
  else{
    this_perf <- performance(prediction(pred, y), measure = "auc")
    auc <- this_perf@y.values[[1]]
  }
  return(auc)
}
```

```{r auc_fGFPCA_nhanes}
auc_mat1 <- df_pred %>%
  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
  select(Y, starts_with("pred"), window) %>%
  group_by(window) %>%
  summarise(auc1 = get_auc(Y, pred480),
            auc2 = get_auc(Y, pred960)) %>%
  filter(window != "[0,480]") %>% 
  select(starts_with("auc"))

colnames(auc_mat1) <- c("0-8am", "0am-4pm")
```

```{r auc_GLMMadaptive_nhanes}
auc_mat2 <- df_pred2 %>%
  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
  select(Y, starts_with("pred"), window) %>%
  group_by(window) %>%
  summarise(auc1 = get_auc(Y, pred480),
            auc2 = get_auc(Y, pred960)) %>%
  filter(window != "[0,480]") %>% 
  select(starts_with("auc"))

colnames(auc_mat2) <- c("0-8am", "0am-4pm")
```

```{r}
data.frame(Window = c("8am-4pm", "4am-12pm"),
           auc_mat1, auc_mat2, check.names = F) %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area Under the ROC curve") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "=1, "fGFPCA" = 2, "GLMMadaptive" = 2)) %>%
  add_header_above(c(" "= 1, "Observed track" = 4))
```

```{r}
nhanes_pred_fgfpca <- df_pred
nhanes_pred_adglmm <- df_pred2
save(nhanes_pred_fgfpca, nhanes_pred_adglmm,
     file = here("Data/ApplOutput.RData"))
```

### 2023-12-04 Docs/manuscript.Rmd
small sample data application

<!-- # NHANES data application -->

>>>>

# NHANES data application

We take 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction.

In the GLMMadaptive model, I could only fit a model with random intercept. It took more than 20 minutes for model fitting.

```{r load_nhanes}
df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
load(here("Data/ApplOutput.RData"))
```

```{r nhance example}
df_nhanes %>% select(SEQN, Z, sind) %>% 
  mutate(Z = factor(Z, levels = 0:1, labels = c("Inactive", "Active"))) %>%
  mutate(SEQN = factor(SEQN, labels = 1:8763)) %>%
  ggplot()+
  geom_tile(aes(x=sind, y = SEQN, fill = Z))+
  labs(x="Time", y="Subject", title = "Overview of NHANES binary activity indicator")+
  theme(axis.text.y = element_blank())

```{r auc_fGFPCA_nhanes}
window3 <- c(0, 480, 960, 1440)

auc_mat_nhanes <- nhanes_pred_fgfpca %>%
  mutate(window = cut(t, breaks = window3, include.lowest = T)) %>% 
  select(Y, starts_with("pred"), window) %>%
  group_by(window) %>%
  summarise(auc1 = get_auc(Y, pred480),
            auc2 = get_auc(Y, pred960)) %>%
  filter(window != "[0,480]") %>% 
  select(starts_with("auc"))

colnames(auc_mat_nhanes) <- c("0-8am", "0am-4pm")
```

```{r auc_GLMMadaptive_nhanes}
auc_mat_nhanes2 <- nhanes_pred_adglmm %>%
  mutate(window = cut(t, breaks = window3, include.lowest = T)) %>% 
  select(Y, starts_with("pred"), window) %>%
  group_by(window) %>%
  summarise(auc1 = get_auc(Y, pred480),
            auc2 = get_auc(Y, pred960)) %>%
  filter(window != "[0,480]") %>% 
  select(starts_with("auc"))

colnames(auc_mat_nhanes2) <- c("0-8am", "0am-4pm")
```

```{r}
data.frame(Window = c("8am-4pm", "4am-12pm"),
           auc_mat_nhanes, auc_mat_nhanes2, check.names = F) %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area Under the ROC curve") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "=1, "fGFPCA" = 2, "GLMMadaptive" = 2)) %>%
  add_header_above(c(" "= 1, "Maximum observation time" = 4))
```

```{r}
rand_id2 <- sample(unique(nhanes_pred_fgfpca$id), 4)

ggarrange(
  nhanes_pred_fgfpca %>% 
    filter(id %in% rand_id2) %>%
    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
    ggplot()+
    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "fGFPCA"),

  nhanes_pred_adglmm %>% 
    filter(id %in% rand_id2) %>%
    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
    ggplot()+
    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "GLMMadaptive"),
  nrow = 1, common.legend = T)
```
